SPIDERMAN: an open-source code to model phase curves and secondary eclipses

### 2017-08-13 ms.tex

Using a geometrical algorithm, the code solves exactly the area of sections of the disc of the planet that are occulted by the star. The code is written in C and optimised to run quickly, with no loss in precision. Approximately 1000 models can be generated per second with typical parameters, making Markov Chain Monte Carlo analyses practicable. The modular nature of the code allows easy comparison of the effect of multiple different brightness distributions of the dataset.

@@

Secondary eclipses and phase curves can give direct insight into the thermal transport processes, and when wavelength resolved also give information on the vertical structure of an exoplanet atmosphere. The details of heat transport in the atmosphere will effect how efficiently energy can be transported between the two hemispheres of the planet.

@@

There appear to be, broadly, two classes of exoplanet atmospheres with regards to heat transport. Hot Jupiters like HD\,189733b with, relatively low temperatures have highly significant hotspot offsets and efficient day-night redistribution \citep{Knutson2007b}. $\mu$ Andromedae, HD\,179949b and HD\,209458b are all significantly hotter than HD\,189733b and display much higher temperature contrasts between the day and night sides, implying inefficient redistribution of energy. \citep{Harrington2006,Cowan2007,zellem2014}. This behavior can be explained with the results of General Circulation Models \citep{Komacek2015,Komacek2016}

@@

\cite{Majeau2012} showed that as longitudinal mapping with phase curves, it is possible to get a two-dimensional map of the dayside of an exoplanet from the secondary eclipse, provided that there is a non-zero impact parameter. This technique works because the inclination of the system results in the occulting limb of the star scanning across the disc of the planet in both longitude and latitude during ingress and egress. The brightness distribution of the planet's permanent dayside will be imprinted on the precise shape of the ingress and egress curves.

@@

@@

\subsection{Analytical integrator}\label{sec:integrator}

@@

\textsc{spiderman} has been written in a modular way, so that users will be able to choose from a variety of models to fit to their data, or provide their own. The currently available models range from fully non-informative to basic physical models. We also provide utilities to allow the results of forward models to be quickly converted into phase curves and eclipses. \textsc{spiderman} projects the model temperature map onto the visible sphere of the planet, and then directly calculates the secondary eclipse and phase curve.

@@

The simplest physical model that one can construct involves defining a bulk planetary temperature or flux, and then a "hotspot" where the size, contrast, and offset from the substellar point in longitude and latitude are model parameters. This model can be used to represent a "two-temperature" planet with different fluxes on the dayside versus the nightside. The hotspot model could also be used to represent the reflection of light due to clouds in optical phase curves, which have been shown to be present and even time varying \citep{Armstrong2016}.

@@

@@

\textsc{spiderman} models the stellar spectrum using a grid of PHOENIX stellar atmospheres calculated by \citet{Husser2013}. Using the user provided wavelength bandpass and instrument response, \textsc{spiderman} calculates the total observed flux for the available models as a function of effective temperature, it then interpolates on this grid using a spline.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{img/system.pdf}
\caption{An example of the graphical output that \textsc{spiderman} can produce. This image is a scale representation of the system, with the planetary flux map included. This is useful for visualising the model. It was generated with the spiderman.plot\_system() command. The image of the star is purely aesthetic.}
\label{fig:plot_system}
\end{center}
\end{figure*}

@@

Planetary limb darkening is a potential sources of uncertainty on the final brightness distribution that would require more detailed atmosphere models to predict. \textsc{spiderman} has the option to include planetary limb darkening in addition to the chosen brightness model with a quadratic law. In principle, measurements of planetary limb darkening can constrain the vertical temperature profile of the planet; however, this would require a full radiative transfer treatment calculation of the atmosphere. For the case of WASP-43b, limb darkening is not expected to be important, since the planet appears to have a near isothermal profile \citep{Stevenson2014}, which should correspond to zero limb darkening. We treat the additional freedom from the limb darkening as a test of the sensitivity to uncertainties of the radial brightness profile of the planet.

@@

Each \emph{visit} displays a decreasing slope, which is modelled as a second order polynomial. The cause of these features is unknown, and they have not been linked to any physical parameters of the telescope \citep{Wakeford2016}. Here, they are also likely blended with the ``breathing" systematics of the instrument, which are caused by the changing thermal environment of the telescope during its orbit changing the focal length of the instrument. The shape of these visit-long effects is slightly different between the three visits used here, so the three parameters of the polynomial are fit seperately for the three visits.

@@

For both models, the MCMC was run for 10 million steps, with the first 200,000 steps discarded as burn-in.

@@

The results of both runs are presented In table \ref{tab:results}. With the exception of $T_n$, which is lower by more than $3\sigma$ in the fit with planetary limb darkening, the physical parameters are consistent. In particular, $\xi$ seems only weakly affected by the presence of limb darkening, the difference being much less than $1\sigma$. This is not unexpected, as $\xi$ effectively parametrises the asymmetry in the flux distribution, while the limb darkening is symmetrical. $\xi$ is directly tied to the details of heat transport in the atmosphere, so the insensitivity of this parameter to limb darkening is useful, as it shows that it can be recovered even with systematic uncertainties in the flux distribution, so long as they are symmetric.

@@

The fit with limb darkening finds a significantly fainter night side flux, suggesting that the nightside flux is an artifact of using a simple physical model. This may be due to a correlation between model parameters making the fit inflexible and the simultaneous fit with the systematics model. Alternatively, the model may be unable to parametrise a feature of the nightside that is blocking the emitted flux, such as clouds.

However, we emphasize that the $\xi$ parameter does not seem to be effected, which raises our confidence in the robustness of the fit to any potential errors in the symmetric radial component of the surface map.

@@

The flat model gave a value of -0.36 - note that this is a good match to the white light value. This gave an acceptable fit, with a chi2 of 15.1 (R chi2 of 1.16) with 14 DOF

>>>>

Using a geometrical algorithm, the code solves exactly the area of sections of the disc of the planet that are occulted by the star. The code is written in C with a user-friendly Python interface, and is optimised to run quickly, with no loss in precision. Approximately 1000 models can be generated per second with typical parameters, making Markov Chain Monte Carlo analyses practicable. The modular nature of the code allows easy comparison of the effect of multiple different brightness distributions of the dataset.

@@

Secondary eclipses and phase curves can give direct insight into the climate of exoplanet atmospheres. The details of heat transport in the atmosphere will effect how efficiently energy can be transported between the two hemispheres of the planet, controlling the day-night contrast, and persistent wind patterns can move the brightest point of the planet from the sub-stellar point. As different wavelengths probe different depths of the atmosphere, spectroscopic phase curves and eclipses also contain information on the vertical temperature structure.

@@

There appear to be two broad classes of exoplanet atmosphere heat circulation. Hot Jupiters like HD\,189733b, with relatively low temperatures have highly significant hotspot offsets and efficient day-night redistribution \citep{Knutson2007b}. $\mu$ Andromedae, HD\,179949b and HD\,209458b are all significantly hotter than HD\,189733b and display much higher temperature contrasts between the day and night sides, implying inefficient redistribution of energy. \citep{Harrington2006,Cowan2007,zellem2014}. This behavior can be explained with the results of General Circulation Models \citep{Komacek2015,Komacek2016}

@@

As well as longitudinal mapping with phase curves, it is possible to get a two-dimensional map of the dayside of an exoplanet from the secondary eclipse, provided that there is a non-zero impact parameter \citep{Rauscher2007}. This technique works because the inclination of the system results in the occulting limb of the star scanning across the disc of the planet in both longitude and latitude during ingress and egress. The brightness distribution of the planet's permanent dayside will be imprinted on the precise shape of the ingress and egress curves. This has been demonstrated for HD\,189733 b by \citep{Majeau2012} ((and others))

@@

In the past, authors studying phase curves have generally used non-physically motivated models to fit the data, such as sine curves ((cite)) or orange slices ((cite)). Whilst these approaches are still useful, \emph{spiderman} enables a \emph{direct} fit to the underlying distribution, and an easy way to compare different models.

@@

\subsection{Integration scheme}\label{sec:integrator}

@@

\textsc{spiderman} has been written in a modular way, so that users will be able to choose from a variety of models to fit to their data, or provide their own. The currently available models range from fully non-informative to basic physical models. We also provide utilities to allow the results of forward models to be quickly converted into phase curves and eclipses. \textsc{spiderman} projects the model temperature or brightness map onto the visible sphere of the planet, and then directly calculates the secondary eclipse and phase curve. The option also exists to include an additional \emph{planetary} limb darkening component to the brightness distribution, which is parameterised as a quadratic law. The major currently available models are listed below.

@@

The simplest physical model that one can construct involves defining a bulk planetary temperature or flux, and then a "hotspot" where the size, contrast, and offset from the substellar point in longitude and latitude are model parameters. This model can be used to represent a "two-temperature" planet with different fluxes on the dayside versus the nightside. The hotspot model could also be used to represent the reflection of light due to clouds in optical phase curves, which have been shown to be present in K2 lightcurves \citep{Demory2013} and are even found to be time varying in HAT-P-7 \citep{Armstrong2016}.

@@

\textsc{spiderman} models the stellar spectrum using a grid of PHOENIX stellar atmospheres calculated by \citet{Husser2013}. It is also possible to use alternative models, or a simple blackbody spectrum, which will typically be sufficient for broadband observations. Using the user provided wavelength bandpass and instrument response, \textsc{spiderman} calculates the total observed flux for the stellar models as a function of effective temperature, it then interpolates on this grid using a spline.

@@

@@

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{img/system.pdf}
\caption{An example of the graphical output that \textsc{spiderman} can produce. This image is a scale representation of the system, with the planetary flux map included. This is useful for visualising the model. It was generated with the spiderman.plot\_system() command. The image of the star is purely aesthetic and is not related to any model parameters (image credit SDO)}
\label{fig:plot_system}
\end{center}
\end{figure*}

@@

Planetary limb darkening is a potential sources of uncertainty on the final brightness distribution that would require more detailed atmosphere models to predict. \textsc{spiderman} has the option to include planetary limb darkening in addition to the chosen brightness model with a quadratic law. In principle, measurements of planetary limb darkening can constrain the vertical temperature profile of the planet; however, this would require a full radiative transfer treatment calculation of the atmosphere. For the case of WASP-43b, where the limb darkening is unknown, we initially fix the limb darkening to zero, and then test the impact of varying the limb darkening, treating the additional freedom as a test of the sensitivity to uncertainties of the radial brightness profile of the planet.

@@

Each \emph{visit} displays a decreasing slope, which is modelled as a second order polynomial. The cause of these features is unknown, and they have not been linked to any physical parameters of the telescope \citep{Wakeford2016}. The shape of these visit-long effects is slightly different between the three visits used here, so the three parameters of the polynomial are fit seperately for the three visits.

@@

For both models, the MCMC was run for 10 million steps, with the first 200,000 steps discarded as burn-in. Convergence was checked by visual inspection of the chains for each model parameter.

@@

The fit was repeated with limb darkening enabled to test the impact of the additional radial flexibility on the model. The results of both runs are presented In table \ref{tab:results}. With the exception of $T_n$, which is lower by more than $3\sigma$ in the fit with planetary limb darkening, the physical parameters are consistent. In particular, $\xi$ seems only weakly affected by the presence of limb darkening, the difference being much less than $1\sigma$. This is not unexpected, as $\xi$ effectively parametrises the asymmetry in the flux distribution, while the limb darkening is symmetrical. $\xi$ is directly tied to the details of heat transport in the atmosphere, so the insensitivity of this parameter to limb darkening is useful, as it shows that it can be recovered even with systematic uncertainties in the flux distribution, so long as they are symmetric.

@@

That the fit with limb darkening enabled finds a significantly fainter night side flux suggests that the nightside flux may be an artifact of using a simple physical model. This may be due to a correlation between model parameters making the fit inflexible and the simultaneous fit with the systematics model. Alternatively, the model may be unable to parametrise a feature of the nightside that is blocking the emitted flux, such as clouds.

@@

We test for the dependence of the value of \xi

The flat model gave a value of -0.36 - which is consistent with the white light value. This gave an acceptable fit, with a chi2 of 15.1 (R chi2 of 1.16) with 14 DOF
